A fast GPU solver for near-Toeplitz tridiagonal systems

### 2015-11-12 paper-draft/paper.tex
updates

Such tridiagonal systems
appear in several numerical schemes
in Computational Fluid Dynamics (CFD)
such as
compact finite differences \cite{lele1992compact},
Alternating Direction Implicit (ADI) methods \cite{1955ADI}, and
numerical solutions of one-dimensional differential equations.
In such applications,
the evaluation of the tridiagonal systems constitutes
a significant portion of the runtime.
Recently, there has been considerable interest in the CFD community
in developing general tridiagonal solvers that
exploit the massive parallelism afforded by
Graphics Processing Units (GPUs)
to speed up the computations
\cite{tutkun2012gpu}
\cite{esfahanian2014efficient}
\cite{GoSt11CR}.
The performance of tridiagonal solvers on the GPU
has been shown to depend on several factors,
such as
choice of algorithm and the associated complexity,
thread utilization,
use of the available memory hierarchy,
and synchronization and control costs
\cite{Zhang2010FTS}.
Many approaches described in the literature
seek to make trade-offs between these factors
to arrive at a solver that best fits the requirements.
But to our knowledge,
the algorithms proposed so far
are applicable to general tridiagonal systems,
and make no attempt to take advantage of
the known structure of the tridiagonal matrix.
The contribution of this paper is twofold:
First, we describe the implementation of a GPU solver
for the specific case of near-Toeplitz systems
that uses the known structure to reduce
computations and memory accesses.
Secondly, we describe the implementation for
multi-GPU clusters using MPI and CUDA,
and apply it to the evaluation of compact finite differences.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[height=250pt]{img/cyclic-reduction.eps}
\end{center}
\caption{Cyclic reduction.}
\label{fig:cyclic-reduction}
\end{figure}

@@

Writing applications that effectively exploit the GPU
requires an understanding of the underlying architecture.
We provide the pertinent details here,
and refer to \cite{GPUcomputingera} for a more complete picture.
Our tests are performed on the NVIDIA Tesla K20 and K40 GPUs,
built on the ``Kepler'' architecture,
an overview of which is available at \cite{Keplerwhitepaper}.

@@

Thread access to global memory is slow,
and is especially inefficient when
successive threads in a block
access locations that are far apart in memory---known
as \emph{uncoalesced} memory access.
Shared memory access can be expected to be
much faster than global memory---however,
the amount of shared memory per SM is limited
(48 KiB for current GPUs).
The amount of shared memory actually used by each block,
can thus limit the number of blocks that
concurrently run on each SM---known as the \emph{occupancy}.
The number of registers per SM is also limited,
and the use of registers can similarly affect occupancy.
Finally, memory transfers between the CPU (host) and GPU (device)
are extremely expensive, limited by the bandwidth of the PCI-e bus.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/gpu-mapping.eps}
\end{center}
\caption{Mapping work to blocks and threads:
systems are mapped to blocks and
indices are mapped to individual threads.}
\label{fig:gpu-mapping}
\end{figure}

@@

In each step of forward reduction,
a thread accesses values from the coefficient arrays
$\bm{a}$, $\bm{b}$ and $\bm{c}$ in a \emph{strided} fashion.
At every subsequent step,
this stride is doubled, while the number of active threads is halved.
In the backward substitution phase,
the strides are halved at each step,
while the number of active threads is doubled.
The long-strided memory accesses towards the end
of the forward reduction
and the beginning of backward substitution
motivates the use of shared memory---strided global memory access is,
in general, associated with performance penalties.
But the limited amount of shared memory places restrictions
on the size of tridiagonal systems that can be solved.
Further,
the memory access pattern in
cyclic reduction leads to \emph{bank conflicts}
\cite{Zhang2010FTS}.
Zhang et al. \cite{Zhang2010FTS} propose a
\emph{hybrid} solver
that uses both cyclic reduction and parallel cyclic reduction
to reduce bank conflicts.
G{\"o}dekke et al. \cite{GoSt11CR}
use a method of separately storing
the even and odd indexed equations
to arrive at a bank-conflict free solver
at the cost of additional shared memory usage.
Shared memory usage is also associated with
lowered GPU occupancy (Section \ref{subsec:gpu-architecture}).
Davidson et al. \cite{davidson2011register}
describe the method of
\emph{register packing}---performing more computations
on \emph{registers}, rather than shared memory---as
a means to reduce shared memory usage in cyclic reduction.
More recently, Esfahanian et al. \cite{esfahanian2014efficient}
describe an implementation of cyclic reduction
that works entirely in global memory,
using additional kernels to reorder data so as to
coalesce memory access.

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/forward-reduction-step.eps}
\end{center}
\caption{Updating $\bm{b}$ in the first forward reduction step.}
\label{fig:forward-reduction-step}
\end{figure}

@@

\begin{itemize}
    \item [] $a_2 = a_3 = a_4 = \hdots \equiv a_0$
    \item [] $b_2 = b_3 = b_4 = \hdots \equiv b_0$
    \item [] $c_2 = c_3 = c_4 = \hdots \equiv c_0$
\end{itemize}
%
we observe that the resulting coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
correspond to the coefficients of a tridiagonal matrix with
exactly the near-Toeplitz structure of $A$.
This form-preserving property of cyclic reduction
has been reported for block Toeplitz tridiagonal systems
by Bini et al.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[height=250pt]{img/cyclic-reduction-precomputing.eps}
\end{center}
\caption{Storing forward reduction coefficients
    from each reduction step $m$.}
\label{fig:cyclic-reduction-precomputing}
\end{figure}

@@

In this section, we describe the implementation of a GPU solver
for solving a given near-Toeplitz tridiagonal system
for $N_{rhs}$ right hand sides.
Our implementation uses the NVIDIA CUDA platform,
We develop two approaches based on the common idea
of precomputed forward reduction coefficients---one that leverages
the GPU's \emph{shared memory},
and the other working directly with global memory.
The GPU kernels for both implementations are relatively
straightforward and compact,
spanning no more than 100 lines.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=400pt]{img/global-and-shared.eps}
\end{center}
\caption{Thread activity in shared memory (top) and global memory (bottom) implementations.}
\label{fig:global-and-shared}
\end{figure}

@@

This implementation works entirely on the GPU's global memory,
Here, we define two kernels - one for the forward reduction step,
and the other for the backward substitution step.
Each kernel is called $log_2(n)-1$ times.
An extra call to the
forward reduction kernel performs
the two-by-two solve.
At each step,
the size of the grid block is determined by the \emph{stride}
between elements accessed at that step.
For the forward reduction phase,
we use $n/2$ threads per block for the first step,
$n/4$ threads for the second step, and so on.
The pattern is reversed for the backward substitution phase,
beginning with 2 threads per block for the first step.
Although this ensures that there are no inactive threads
at any stage,
the occupancy of the GPU is still very low
during the end of forward reduction
and the end of backward substitution.

@@

In the shared memory approach (Fig. \ref{fig:global-and-shared}),
we launch a single kernel to perform the entire
cyclic reduction solve.
The kernel is launched with $n/2$ threads per block.
Each thread block allocated a block of shared memory of size $n/2$.
Each thread performs the first reduction step
[Eq. (\ref{eqn:precomputed-forward-reduction-step})]
by accessing the required values
$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
storing the result in shared memory.
In subsequent reduction steps,
$d_i$, $d_{i-1}$ and $d_{i+1}$
are accessed from shared memory,
avoiding the uncoalesced global memory accesses
seen in the global memory implementation.
In each back substitution step,
threads overwrite the existing values in shared memory
with the values of the solution.
In the final step,
shared memory is filled completely
with the even-indexed solution values.
Each thread then computes an odd-indexed solution value,
storing it directly in global memory
and copies the even-indexed solution value
from shared memory to global memory.
The entire solution is done in a single kernel launch
to hide the latency of
memory transfers between global and shared memory.
Explicit synchronization between the threads of a block
is required within the kernel at the end of each step.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/bench-2d.eps}
\caption{Relative solver performance for 2-D problems. Relative time defined as:}
$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
\label{fig:bench-2d}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/bench-3d.eps}
\caption{Relative solver performance for 3-D problems. Relative time defined as:}
$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
\label{fig:bench-3d}
\end{center}
\end{figure}

@@

The CPU code is compiled with the Intel C compiler (version 15.0),
and run (with OpenMP support) on up to
16 independent cores of the the same shared-memory node
(one thread per core).
The CPU is an
Intel Xeon Processor E5-2670 v2 (2.50 GHz, 25 MB Smart Cache).
GPU code is compiled with the CUDA toolkit (version 6.5.14),
and run on the
NVIDIA Tesla K20 Accelerator.
When measuring GPU performance (both CUSPARSE and NEATO),
we do \emph{not} include the cost
of data transfer between the CPU and GPU.
\texttt{-O2} level compiler optimizations are turned on for both
CPU and GPU code;
no further optimization options are enabled in either case.
Of course, we use double precision for all solvers.
All timings reported are for the tridiagonal solution
and are averaged over several (100) solves.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
\caption{Comparison of global memory and shared
    memory implementations of NEATO (2D problems).}
\label{fig:global-vs-shared-2d}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-3d.eps}
\caption{Comparison of global memory and shared memory
    implementations of NEATO (3D problems).}
\label{fig:global-vs-shared-3d}
\end{center}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/computational-domain.eps}
\caption{Computational domain in 3-D.}
\label{fig:computational-domain}
\end{center}
\end{figure}

@@

\begin{equation}
    f^{\prime}_{n} + 2f^{\prime}_{n-1s}
    =
    \frac{5f_{n} - 4f_{n-2} 1 f_{n-2}}{dx}
\end{equation}
%
This boundary treatment maintains the tridiagonal identity
of the system to be solved, which then has the near-Toeplitz form:

\begin{equation} \label{eqn:compact-tridiagonal-system}
 \begin{bmatrix}
     1&2\\
     1/4&1&1/4\\
     &1/4&1&1/4\\
     &&1/4&1&1/4\\
     &&&1/4&1&1/4\\
     &&&&&\ddots\\
     &&&&&&\ddots\\
     &&&&&&&\ddots\\
     &&&&&&&2&1
  \end{bmatrix}
  \begin{bmatrix}
      f^{\prime}_1 \\
      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     -5f_1 + 4f_2 + f_3\\
     f_{3} - f_{1}\\
     f_{4} - f_{2}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     f_{n} - f_{n-2}\\
     5f_{n} - 4f_{n-1} - f_{n-2}
  \end{bmatrix}
\end{equation}

@@

The computational domain we consider
(Fig. \ref{fig:computational-domain})
is a structured grid comprised of
$nx \times ny \times nz$ grid points.
While evaluating the derivative in, say, the x-direction,
the computational domain is thought of as
a collection of \emph{grid lines}---each consisting of
$nx$ grid points---oriented in the x-direction.
For each grid line,
we may assemble a tridiagonal system of size $nx$, as in
Eq. (\ref{eqn:compact-tridiagonal-system}).
Solving for the all derivatives in the $x$ direction then involves
solving this tridiagonal system for $ny \times nz$ right hand sides.
On a single GPU,
this is done easily using the NEATO solver.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=350pt]{img/subdomains-grid-lines.eps}
\caption{A line of subdomains and a single grid line composed of ``local'' grid lines.
    Each subdomain is mapped to a single GPU, or \emph{process},
    and each process is identified by its ``rank''.}
\label{fig:subdomains-grid-lines}
\end{center}
\end{figure}

@@

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
x_{r,1}^p \\
x_{r,2}^p \\
x_{r,3}^p \\
x_{r,4}^p \\
\vdots \\
x_{r,m}^p
\end{bmatrix}
=
\begin{bmatrix}
r_1^p \\
r_2^p \\
r_3^p \\
r_4^p \\
\vdots \\
r_m^p
\end{bmatrix} & \label{eqn:primary-system-expanded} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
u_1^p \\
u_2^p \\
u_3^p \\
u_4^p \\
\vdots \\
u_m^p
\end{bmatrix}
=
\begin{bmatrix}
-a_1^p \\
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix} & \label{eqn:secondary-system-1-expanded} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
l_1^p \\
l_2^p \\
l_3^p \\
l_4^p \\
\vdots \\
l_m^p
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
\vdots \\
-c_m^p
\end{bmatrix} & \label{eqn:secondary-system-2-expanded}
\end{align}
%
We refer to the subsystem in Eq. (\ref{eqn:primary-system-expanded})
the ``primary'' system, and the subsystems in
Eqs. (\ref{eqn:secondary-system-1-expanded}) and
(\ref{eqn:secondary-system-2-expanded})
the ``secondary'' systems.
Additionally, the following ``reduced'' system is solved:

@@

We present a specialization of the above algorithm
applied to the distributed near-Toeplitz systems that
appear in compact finite difference evaluations,
and adapted for GPUs.
Our algorithm is outlined in \ref{fig:compact-algorithm},
and is described for the case of evaluating derivatives
in the direction along which successive
function values in a subdomain are stored contigiously in memory,
i.e., the ``fastest'' coordinate direction.
Each step of the algorithm must be implemented on the GPU
to avoid data transfer to and from the CPU,
which is prohibitively expensive for large problems.
Thus, we have several kernels to implement the algorithm.
For communication of data between processes,
we use the Message Passing Interface (MPI),
and leverage the NVIDIA GPUDirect Technology
for GPU-GPU communication.
The purpose of each CUDA kernel and MPI call used
is described in Table \ref{table:compact-algorithm}.

\begin{figure}[h!]
\begin{center}
\includegraphics[trim={0 7cm 0 7cm},clip,width=500pt]{img/compact-algorithm.pdf}
\centering
\caption{Algorithm for evaluating compact finite differences on multiple GPUs,
(right: CUDA kernels and MPI calls used)}
\label{fig:compact-algorithm}
\end{center}
\end{figure}

>>>>

Such tridiagonal systems
appear in several numerical schemes
in Computational Fluid Dynamics (CFD)
such as
compact finite differences
\cite{lele1992compact,kennedy1994several},
Alternating Direction Implicit (ADI) methods \cite{1955ADI}, and
numerical solutions of one-dimensional differential equations.
In such applications,
the evaluation of the tridiagonal systems constitutes
a significant portion of the runtime.
Recently, there has been considerable interest in the CFD community
in developing general tridiagonal solvers that
exploit the massive parallelism afforded by
Graphics Processing Units (GPUs)
to speed up the computations
\cite{tutkun2012gpu,esfahanian2014efficient,GoSt11CR}.
The performance of tridiagonal solvers on the GPU
has been shown to depend on several factors,
such as
choice of algorithm and the associated complexity,
thread utilization,
use of the available memory hierarchy,
and synchronization and control costs
\cite{Zhang2010FTS}.
Many approaches described in the literature
seek to make trade-offs between these factors
to arrive at a solver that best fits the requirements.
But to our knowledge,
the algorithms proposed so far
are applicable to general tridiagonal systems,
and make no attempt to take advantage of
the known structure of the tridiagonal matrix.
The contribution of this paper is twofold:
First, we describe the implementation of a GPU solver
for the specific case of near-Toeplitz systems
that uses the known structure to reduce
computations and memory accesses.
Secondly, we describe the implementation for
multi-GPU clusters using MPI and CUDA,
and apply it to the evaluation of compact finite differences.

@@

\begin{figure}
\begin{center}
\includegraphics[height=250pt]{img/cyclic-reduction.eps}
\end{center}
\caption{Cyclic reduction.}
\label{fig:cyclic-reduction}
\end{figure}

@@

Writing applications that effectively exploit the GPU
requires an understanding of the underlying architecture.
We provide the pertinent details here
and refer to \cite{GPUcomputingera} for a more complete picture.
Our tests are performed on the NVIDIA Tesla K20 and K40 GPUs,
built on the ``Kepler'' architecture,
an overview of which is available at \cite{Keplerwhitepaper}.

@@

Thread access to global memory is slow,
and is especially inefficient when
successive threads in a block
access locations that are far apart in memory---known
as \emph{uncoalesced} memory access.
Shared memory access can be expected to be
much faster than global memory---however,
the amount of shared memory per SM is limited
(48 KiB for current GPUs).
The amount of shared memory actually used by each block
can thus limit the number of blocks that
concurrently run on each SM---known as the \emph{occupancy}.
The number of registers per SM is also limited,
and the use of registers can similarly affect occupancy.
Finally, memory transfers between the CPU (host) and GPU (device)
are extremely expensive, limited by the bandwidth of the PCI-e bus.

@@

\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/gpu-mapping.eps}
\end{center}
\caption{Mapping work to blocks and threads:
systems are mapped to blocks and
indices are mapped to individual threads.}
\label{fig:gpu-mapping}
\end{figure}

@@

In each step of forward reduction,
a thread accesses values from the coefficient arrays
$\bm{a}$, $\bm{b}$ and $\bm{c}$ in a \emph{strided} fashion.
At every subsequent step,
this stride is doubled, while the number of active threads is halved.
In the backward substitution phase,
the strides are halved at each step,
while the number of active threads is doubled.
The long-strided memory accesses towards the end
of the forward reduction
and the beginning of backward substitution
motivates the use of shared memory---strided global memory access is,
in general, associated with performance penalties.
But the limited amount of shared memory places restrictions
on the size of tridiagonal systems that can be solved.
Further,
the memory access pattern in
cyclic reduction leads to \emph{bank conflicts}
\cite{Zhang2010FTS}.

Much work has been done on optimizing cyclic reduction performance
ong GPUs.
Zhang et al. \cite{Zhang2010FTS} propose a
\emph{hybrid} solver
that uses both cyclic reduction and parallel cyclic reduction
to reduce bank conflicts.
G{\"o}dekke et al. \cite{GoSt11CR}
use a method of separately storing
the even and odd indexed equations
to arrive at a bank-conflict free solver
at the cost of additional shared memory usage.
Shared memory usage is also associated with
lowered GPU occupancy (Section \ref{subsec:gpu-architecture}).
Davidson et al. \cite{davidson2011register}
describe the method of
\emph{register packing}---performing more computations
on \emph{registers}, rather than shared memory---as
a means to reduce shared memory usage in cyclic reduction.
Esfahanian et. al, \cite{esfahanian2014efficient}
avoid shared memory (and associated bank conflicts entirely)
using a data rearrangement scheme to improve global memory access.
Our approach takes a different route,
and is focussed on near-Toeplitz systems---exploiting
the relatively simple matrix structure
to reduce the number of computations and memory accesses
performed by each thread
at every cyclic reduction step.

\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/forward-reduction-step.eps}
\end{center}
\caption{Updating $\bm{b}$ in the first forward reduction step.}
\label{fig:forward-reduction-step}
\end{figure}

@@

\begin{itemize}
    \item [] $a_2 = a_3 = a_4 = \hdots \equiv a_0$
    \item [] $b_2 = b_3 = b_4 = \hdots \equiv b_0$
    \item [] $c_2 = c_3 = c_4 = \hdots \equiv c_0$
\end{itemize}
%
we observe that the resulting coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
correspond to the coefficients of a tridiagonal matrix with
exactly the near-Toeplitz structure of $A$.
This form-preserving property of cyclic reduction
has been reported for block Toeplitz tridiagonal systems
by Bini et al \cite{bini}.

@@

\begin{figure}
\begin{center}
\includegraphics[height=250pt]{img/cyclic-reduction-precomputing.eps}
\end{center}
\caption{Storing forward reduction coefficients
    from each reduction step $m$.}
\label{fig:cyclic-reduction-precomputing}
\end{figure}

@@

In this section, we describe the implementation of a GPU solver
for solving a given near-Toeplitz tridiagonal system
for $N_{rhs}$ right hand sides.
Our implementation uses the NVIDIA CUDA platform for writing GPU kernels,
interfaced via the Python library PyCUDA \cite{kloeckner_pycuda_2012}.
We develop two approaches based on the common idea
of precomputed forward reduction coefficients---one that leverages
the GPU's \emph{shared memory},
and the other working directly with global memory.
The GPU kernels for both implementations are relatively
straightforward and compact,
spanning no more than 100 lines of code.

@@

\begin{figure}
\begin{center}
\includegraphics[width=400pt]{img/global-and-shared.eps}
\end{center}
\caption{Thread activity in shared memory (top) and global memory (bottom) implementations.}
\label{fig:global-and-shared}
\end{figure}

@@

This implementation works entirely on the GPU's global memory,
Here, we define two kernels - one for the forward reduction step,
and the other for the backward substitution step.
Each kernel is called $log_2(n)-1$ times.
An extra call to the
forward reduction kernel performs
the two-by-two solve.
At each step,
the size of the thread blocks is determined by the \emph{stride}
between elements accessed at that step.
For the forward reduction phase,
we use $n/2$ threads per block for the first step,
$n/4$ threads for the second step, and so on.
The pattern is reversed for the backward substitution phase,
beginning with 2 threads per block for the first step.
Although this ensures that there are no inactive threads
at any stage,
the occupancy of the GPU is still very low
during the end of forward reduction
and the end of backward substitution.

@@

In the shared memory approach (Fig. \ref{fig:global-and-shared}),
we launch a single kernel to perform the entire
cyclic reduction solve.
The kernel is launched with $n/2$ threads per block.
Each thread block is allocated a block of shared memory of size $n/2$.
Each thread of a block performs the first reduction step
[Eq. (\ref{eqn:precomputed-forward-reduction-step})]
by accessing the required values
$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
storing the result in shared memory.
In subsequent reduction steps,
$d_i$, $d_{i-1}$ and $d_{i+1}$
are accessed from shared memory,
avoiding the uncoalesced global memory accesses
seen in the global memory implementation.
In each back substitution step,
threads overwrite the existing values in shared memory
with the values of the solution.
In the final step,
shared memory is filled completely
with the even-indexed solution values.
Each thread then computes an odd-indexed solution value,
storing it directly in global memory
and copies the even-indexed solution value
from shared memory to global memory.
The entire solution is done in a single kernel launch
to hide the latency of
memory transfers between global and shared memory.
Explicit synchronization between the threads of a block
is required within the kernel at the end of each step.

@@

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/bench-2d.eps}
\caption{Relative solver performance for 2-D problems. Relative time defined as:}
$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
\label{fig:bench-2d}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/bench-3d.eps}
\caption{Relative solver performance for 3-D problems. Relative time defined as:}
$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
\label{fig:bench-3d}
\end{center}
\end{figure}

@@

The CPU code is compiled with the Intel C compiler (version 15.0),
and run (with OpenMP support) on up to
16 independent cores of the the same shared-memory node
(one thread per core).
The CPU is an
Intel Xeon Processor E5-2670 v2 (2.50 GHz, 25 MB Smart Cache).
GPU code is compiled with the CUDA toolkit (version 6.5.14),
and run on the
NVIDIA Tesla K20 Accelerator.
When measuring GPU performance (both CUSPARSE and NEATO),
we do \emph{not} include the cost
of data transfer between the CPU and GPU.
\texttt{-O2} level compiler optimizations are turned on for both
CPU and GPU code;
no further optimization options are enabled in either case.
Of course, we use double precision for all solvers.
The timings reported are
kernel \emph{execution} times, i.e.,
the time for all kernel(s) to execute completely
before returning the program control to the CPU.
All timings are averaged over 100 tridiagonal solves.

@@

\begin{figure}
\begin{center}
\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
\caption{Comparison of global memory and shared
    memory implementations of NEATO (2D problems).}
\label{fig:global-vs-shared-2d}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-3d.eps}
\caption{Comparison of global memory and shared memory
    implementations of NEATO (3D problems).}
\label{fig:global-vs-shared-3d}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[height=150pt]{img/computational-domain.eps}
\caption{Computational domain in 3-D.}
\label{fig:computational-domain}
\end{center}
\end{figure}

@@

\begin{equation}
    f^{\prime}_{n} + 2f^{\prime}_{n-1}
    =
    \frac{5f_{n} - 4f_{n-2} -  f_{n-1}}{dx}
\end{equation}
%
This boundary treatment maintains the tridiagonal identity
of the system to be solved, which then has the near-Toeplitz form:

\begin{equation} \label{eqn:compact-tridiagonal-system}
 \begin{bmatrix}
     1&2\\
     1/4&1&1/4\\
     &1/4&1&1/4\\
     &&1/4&1&1/4\\
     &&&1/4&1&1/4\\
     &&&&&\ddots\\
     &&&&&&\ddots\\
     &&&&&&&\ddots\\
     &&&&&&&2&1
  \end{bmatrix}
  \begin{bmatrix}
      f^{\prime}_1 \\
      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     \frac{-5f_1 + 4f_2 + f_3}{2}\\
     \frac{3(f_{3} - f_{1})}{4}\\
     \frac{3(f_{4} - f_{2})}{4}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     \frac{3(f_{n} - f_{n-2})}{4}\\
     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2}
  \end{bmatrix}
\end{equation}

@@

The computational domain we consider
(Fig. \ref{fig:computational-domain})
is a structured grid comprised of
$nx \times ny \times nz$ grid points.
While evaluating the derivative in, say, the x-direction,
the computational domain is thought of as
a collection of \emph{grid lines}---each consisting of
$nx$ grid points---oriented in the x-direction.
For each grid line,
we may assemble a tridiagonal system of size $nx$, as in
Eq. (\ref{eqn:compact-tridiagonal-system}).
Solving for the all derivatives in the $x$ direction then involves
solving this tridiagonal system for $ny \times nz$ right hand sides.
Some work has been done on the solution of these
tridiagonal systems using GPUs---Tutkun et. al
\cite{tutkun2012gpu} compute the inverse of the coefficient matrix,
and use the GPU to multiply the inverse with the right-hand sides.
Wei et al. \cite{wei2013parallelizing}
use the parallel cyclic reduction (PCR) algorithm
to solve similar systems appearing in
the 2-D ADI problem.
More recently, Esfahanian et. al \cite{esfahanian2014efficient}
use a variant of cyclic reduction with
data rearrangements kernels for better global memory access patterns.
Our approach uses the NEATO solver described above
to solve for the right hand sides at each grid line,
which takes advantage of the known near-Toepltiz structure
of the coefficient matrix arising
in the compact finite difference scheme.

@@

\begin{figure}
\begin{center}
\includegraphics[width=350pt]{img/subdomains-grid-lines.eps}
\caption{A line of subdomains and a single grid line composed of ``local'' grid lines.
    Each subdomain is mapped to a single GPU, or \emph{process},
    and each process is identified by its ``rank''.}
\label{fig:subdomains-grid-lines}
\end{center}
\end{figure}

@@

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
x_{r,1}^p \\
x_{r,2}^p \\
x_{r,3}^p \\
x_{r,4}^p \\
\vdots \\
x_{r,m}^p
\end{bmatrix}
=
\begin{bmatrix}
r_1^p \\
r_2^p \\
r_3^p \\
r_4^p \\
\vdots \\
r_m^p
\end{bmatrix} & \label{eqn:primary-system-expanded} \\
\end{align}

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
u_1^p \\
u_2^p \\
u_3^p \\
u_4^p \\
\vdots \\
u_m^p
\end{bmatrix}
=
\begin{bmatrix}
-a_1^p \\
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix} & \label{eqn:secondary-system-1-expanded} \\
\end{align}
%
\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
l_1^p \\
l_2^p \\
l_3^p \\
l_4^p \\
\vdots \\
l_m^p
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
\vdots \\
-c_m^p
\end{bmatrix} & \label{eqn:secondary-system-2-expanded}
\end{align}
%
We refer to the subsystem in Eq. (\ref{eqn:primary-system-expanded})
the ``primary'' system, and the subsystems in
Eqs. (\ref{eqn:secondary-system-1-expanded}) and
(\ref{eqn:secondary-system-2-expanded})
the ``secondary'' systems.
Additionally, the following ``reduced'' system is solved:

@@

We present a specialization of the above algorithm
applied to the distributed near-Toeplitz systems that
appear in compact finite difference evaluations,
and adapted for GPUs.
Our algorithm is outlined in \ref{fig:compact-algorithm},
and is described for the case of evaluating derivatives
in the direction along which successive
function values in a subdomain are stored contigiously in memory,
i.e., the ``fastest'' coordinate direction.
Each step of the algorithm must be implemented on the GPU
to avoid data transfer to and from the CPU,
which is prohibitively expensive for large problems.
Thus, we have several kernels to implement the algorithm.
For communication of data between processes,
we use the Message Passing Interface (MPI),
and leverage the NVIDIA GPUDirect Technology
for GPU-GPU communication.
MPI is interfaced via the mpi4py \cite{dalcin2005mpi}
Python library.
The purpose of each CUDA kernel and MPI call used
is described in Table \ref{table:compact-algorithm}.

\begin{figure}
\begin{center}
\includegraphics[trim={0 7cm 0 7cm},clip,width=500pt]{img/compact-algorithm.pdf}
\centering
\caption{Algorithm for evaluating compact finite differences on multiple GPUs,
(right: CUDA kernels and MPI calls used)}
\label{fig:compact-algorithm}
\end{center}
\end{figure}

\begin{table}[]
\centering
\caption{Purpose of kernels and MPI calls in compact finite difference application}
\label{table:compact-algorithm}
\begin{tabular}{|l|p{6cm}|}
\hline
Memcpy3d            & Copy noncontiguous boundary information
                      of the function values
                      to and from contiguous halo arrays \\ \hline
ISend, IRecv        & Perform halo swaps with i-1 and i+1 processes \\ \hline
computeRHSKernel    & Apply pointwise stencil operator to compute
                      RHS at each grid point in the subdomain,
                      using the halo values near the boundaries\\ \hline
NEATO               & Solve the primary near-Toeplitz tridiagonal systems
                      for the computed right hand sides, giving $x_r$ \\ \hline
copyFacesKernel     & Copy the left and right faces of $x_r$
                      into a single contiguous array \\ \hline
MPI\_Gather         & Gather the data required to assemble
                      the reduced system at rank 0 \\ \hline
reducedSolverKernel & Solve the reduced systems for parameters
                      $\alpha^p$ and $\beta^p$
                      for each grid line \\ \hline
MPI\_Scatter        & Scatter the parameters 
                      $\alpha^p$ and $\beta^p$
                      from rank 0 to all the processes \\ \hline
sumSolutionsKernel  & Sum the primary and secondary solutions
                      to compute the local part of the solution \\ \hline
\end{tabular}
\end{table}

\begin{figure}
\begin{center}
\includegraphics[width=200pt]{img/constructing-reduced-system.eps}
\caption{Construction of the reduced system and scattering of parameters.}
\label{fig:constructing-reduced-system}
\end{center}
\end{figure}
